GMSE: an R package for generalised management strategy evaluation

### 2018-04-26 notebook/ms/SI4.Rmd
Small changes

| Parameter         | Value    | Description                                      |
|-------------------|----------|--------------------------------------------------|
| `remove_pr`       | 0.122    | Goose density-independent mortality probability  |
| `lambda`          | 0.275    | Expected offspring production per time step      |
| `res_death_K`     | 93870    | Goose carrying capacity (on adult mortality)     |
| `RESOURCE_ini`    | 35000    | Initial goose abundance                          |
| `manage_target`   | 70000    | Manager's target goose abundance                 |
| `res_death_type`  | 3        | Mortality (density and density-indepent sources) |

@@

Additionally, we continue to use the following values for consistency, except in the case of `stakeholders`, where we reduce the number of farmers to `stakeholders = 8` and increase `user_budget` to 100000. This is done to for two reasons. First, it speeds up simulations for the purpose of demonstration; second, it makes the presentation of landscape ownership easier (see below).

| Parameter         | Value    | Description                                      |
|-------------------|----------|--------------------------------------------------|
| `manager_budget`  | 10000    | Manager's budget for setting policy options      |
| `user_budget`     | 100000   | Users' budgets for actions                       |
| `public_land`     | 0.4      | Proportion of the landscape that is public       |
| `stakeholders`    | 8        | Number of stakeholders                           |
| `land_ownership`  | TRUE     | Users own landscape cells                        |
| `res_consume`     | 0.02     | Landscape cell output consumed by a resource     | 
| `observe_type`    | 3        | Observation model type (survey)                  |
| `agent_view`      | 1        | Cells managers can see when conducting a survey  |

@@

We now recreate the simulations in [Supporting Information 3](), which were run using the `gmse` function, in `gmse_apply`. Doing so requires us to first initialise simulations using one run of `gmse_apply`, then loop through multiple time steps that again call `gmse_apply` and saving the results of interest. Following instructions in [Supporting Information 1](), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.

```{r, echo = TRUE, eval = TRUE}
sim_old  <- gmse_apply(get_res = "Full", remove_pr = 0.122, lambda = 0.275,
                       res_death_K = 93870, RESOURCE_ini = 35000, 
                       manage_target = 70000, res_death_type = 3,
                       manager_budget = 10000, user_budget = 100000,
                       public_land = 0.4, stakeholders = 8, 
                       land_ownership = TRUE, res_consume = 0.02, 
                       observe_type = 3, agent_view = 1);
```

@@

The above output from `sim_sum_1` shows the table that holds the information we were interested in pulling out of our simulation results. All of this information was available under the list element `sim_new$basic_output`, but other list elements of `sim_new` might also be useful to record. It is important to also remember that this example of `gmse_apply` is using the default resource, observation, manager, and user submodels. Custom submodels could produce different outputs in `sim_new` (see [Supporting Information 1]() for examples). For default options, there are some list elements that might be especially useful. All of these elements can be edited *within the above loop* to dynamically adjust simulations.

- `sim_new$resource_array`: A table holding all information on resources. Rows correspond to a unique resource, and columns correspond to resource properties: (1) ID, (2-4) resource types (not currently in use), (5) x-location, (6) y-location, (7) movement parameter, (8) time, (9) density independent mortality parameter (`remove_pr`), (10) reproduction parameter (`lambda`), (11) offspring number, (12) age, (13-14) observation columns, (15) consumption rate (`res_consume`), and (16-20) recorded experiences of user actions.
- `sim_new$AGENTS`: A table holding basic information on agents (manager and users). Rows correspond to a unique agent, and columns correspond to agent properties: (1) ID, (2) type (0 for the manager, 1 for users), (3-4) additional type options not currently in use, (5-6), x and y locations (usually ignored), (7) movement parameter (usually ignored), (8) time, (9) agent's viewing ability in cells (`agent_view`), (10) error parameter, (11-12) values for holding marks and tallies of resources, (13-15) values for holding observations, (16) yield from landscape cells, (17) budget (`manager_budget` and `user_budget`).
- `sim_new$observation_vector`: Estimate of total resource number from the observation model (`observation_array` also holds this information in a different way depending on `observe_type`)
- `sim_new$LAND`: The landscape on which interactions occur, which is stored as a 3D array with `land_dim_1` rows, `land_dim_2` columns, and 3 layers. Layer 1 (`sim_new$LAND[,,1]`) is not used, but could be used to store values that affect resources and agents. Layer 2 (`sim_new$LAND[,,2]`) stores crop yield from a cell, and layer 3 (`sim_new$LAND[,,3]`) stores the owner of the cell (value corresponds to the agent's ID).
- `sim_new$manage_vector`: The cost of each action as set by the manager. For even more fine-tuning, individual costs for actions can be set for each user in `sim_new$manager_array`.
- `sim_new$user_vector`: The total number of actions performed by each user. A more detailed breakdown of actions by individual users is held in `sim_new$user_array`.

We now show how to adjust the landscape to manually adjust land ownership in `gmse_apply`.

@@

By default, all farmers in GMSE are allocated the same number of landscape cells, which are simply placed in order of the farmer's ID. Public landscape is produced by placing landscape cells that are technically owned by the manager, and therefore have landscape cell values of 0. The image below shows this landscape for the eight farmers from `sim_old`.

@@

```{r, echo = TRUE, eval = TRUE}
sim_old  <- gmse_apply(get_res = "Full", remove_pr = 0.122, lambda = 0.275,
                       res_death_K = 93870, RESOURCE_ini = 35000, 
                       manage_target = 70000, res_death_type = 3,
                       manager_budget = 10000, user_budget = 100000,
                       public_land = 0.4, stakeholders = 8, 
                       land_ownership = TRUE, res_consume = 0.02, 
                       observe_type = 3, agent_view = 1);
```

@@

Perhaps we want to assume that farmers have different budgets, which are perhaps correlated in some way to the number of landscape cells that they own. Custom user budgets can be set by manipulating `sim_old$AGENTS`, the last column of which holds the budget for each user. Agent IDs (as stored on the landscape above) correspond to rows of `sim_old$AGENTS`, so individual budgets can be directly input as desired. We can do this manually (e.g., `sim_old$AGENTS[2, 17] <- 4000`), or, alternatively, if farmer budget positively correlates to landscape owned, we can use a loop to input values as below.

@@

```{r, echo = FALSE}
budget_table <- rbind(sim_old$AGENTS[,1], sim_old$AGENTS[,17]);
row.names(budget_table) <- c("ID", "Budget");
kable(budget_table, caption = "Custom created budgets for manager and users");
```

@@

Lastly, we consider a more nuanced change to simulations, in which the rules for movement of resources are modified to account for density-dependence. Assume that geese tend to avoid aggregating, such that if a goose is located on the same cell as too many other geese, then it will move with a high probability at the end of a time step. Programming this movement rule can be accomplished by creating a new function to apply to the resource data array `sim_old$resource_array`. Below, a custom function is defined that causes a goose to move up to 5 cells in any direction if it finds itself on a cell with more than 10 other geese. As with default GMSE simulations, movement is based on a torus landscape.

```{r}
avoid_aggregation <- function(goose_table, land_dim_1 = 100, land_dim_2 = 100){
    goose_number <- dim(goose_table)[1]    # How many geese are there?
    for(goose in 1:goose_number){          # Loop through all rows of geese
        x_loc  <- goose_table[goose, 5];
        y_loc  <- goose_table[goose, 6];
        shared <- sum(goose_table[,5] == x_loc & goose_table[,6] == y_loc);
        if(shared > 10){
            new_x <- x_loc <- sample(x = -5:5, size = 1);
            new_y <- y_loc <- sample(x = -5:5, size = 1);
            if(new_x < 0){ # The 'if' statements below apply the torus
                new_x <- land_dim_1 + new_x;
            }
            if(new_x >= land_dim_1){
                new_x <- new_x - land_dim_1;
            }
            if(new_y < 0){
                new_y <- land_dim_2 + new_x;
            }
            if(new_y >= land_dim_2){
                new_y <- new_y - land_dim_2;
            }
            goose_table[goose, 5] <- new_x;
            goose_table[goose, 6] <- new_y;
        }
    }
    return(goose_table);
}
```

>>>>

| Parameter         | Value    | Description                                         |
|-------------------|----------|-----------------------------------------------------|
| `remove_pr`       | 0.122    | Goose density-independent mortality probability     |
| `lambda`          | 0.275    | Expected offspring production per time step         |
| `res_death_K`     | 93870    | Goose carrying capacity (on adult mortality)        |
| `RESOURCE_ini`    | 35000    | Initial goose abundance                             |
| `manage_target`   | 70000    | Manager's target goose abundance                    |
| `res_death_type`  | 3        | Mortality (density and density-independent sources) |

@@

Additionally, we continue to use the following values for consistency, except in the case of `stakeholders`, where we reduce the number of farmers to `stakeholders = 8`. This is done to for two reasons. First, it speeds up simulations for the purpose of demonstration; second, it makes the presentation of landscape ownership easier (see below).

| Parameter         | Value    | Description                                      |
|-------------------|----------|--------------------------------------------------|
| `manager_budget`  | 10000    | Manager's budget for setting policy options      |
| `user_budget`     | 10000    | Users' budgets for actions                       |
| `public_land`     | 0.4      | Proportion of the landscape that is public       |
| `stakeholders`    | 8        | Number of stakeholders                           |
| `land_ownership`  | TRUE     | Users own landscape cells                        |
| `res_consume`     | 0.02     | Landscape cell output consumed by a resource     | 
| `observe_type`    | 3        | Observation model type (survey)                  |
| `agent_view`      | 1        | Cells managers can see when conducting a survey  |

@@

We now recreate the simulations in [Supporting Information 3](), which were run using the `gmse` function, in `gmse_apply`. Doing so requires us to first initialise simulations using one call of `gmse_apply`, then loop through multiple time steps that again call `gmse_apply`; results of interest are recorded in a data frame (`sim_sum_1`). Following the protocol introduced in [Supporting Information 2](), we can call the initialising simulation `sim_old`, and use the code below to read in the relevant parameter values.

```{r, echo = TRUE, eval = TRUE}
sim_old  <- gmse_apply(get_res = "Full", remove_pr = 0.122, lambda = 0.275,
                       res_death_K = 93870, RESOURCE_ini = 35000, 
                       manage_target = 70000, res_death_type = 3,
                       manager_budget = 10000, user_budget = 100000,
                       public_land = 0.4, stakeholders = 8, res_consume = 0.02, 
                       res_birth_K = 200000, land_ownership = TRUE,
                       observe_type = 3, agent_view = 1, converge_crit = 0.01,
                       ga_mingen = 200);
```

@@

The above output from `sim_sum_1` shows the data frame that holds the information we were interested in pulling out of our simulation results. All of this information was available under the list element `sim_new$basic_output`, but other list elements of `sim_new` might also be useful to record. It is important to remember that this example of `gmse_apply` is using the default resource, observation, manager, and user submodels. Custom submodels could produce different outputs in `sim_new` (see [Supporting Information 2]() for examples). For default submodels, there are some list elements that might be especially useful. These elements can potentially be edited *within the above loop* to dynamically adjust simulations.

- `sim_new$resource_array`: A table holding all information on resources. Rows correspond to discrete resources, and columns correspond to resource properties: (1) ID, (2-4) types (not currently in use), (5) x-location, (6) y-location, (7) movement parameter, (8) time, (9) density independent mortality parameter (`remove_pr`), (10) reproduction parameter (`lambda`), (11) offspring number, (12) age, (13-14) observation columns, (15) consumption rate (`res_consume`), and (16-20) recorded experiences of user actions (e.g., was the resource culled or scared?).
- `sim_new$AGENTS`: A table holding basic information on agents (manager and users). Rows correspond to a unique agent, and columns correspond to agent properties: (1) ID, (2) type (0 for the manager, 1 for users), (3-4) additional type options not currently in use, (5-6), x and y locations (usually ignored), (7) movement parameter (usually ignored), (8) time, (9) agent's viewing ability in cells (`agent_view`), (10) error parameter, (11-12) values for holding marks and tallies of resources, (13-15) values for holding observations, (16) yield from landscape cells, (17) budget (`manager_budget` and `user_budget`).
- `sim_new$observation_vector`: Estimate of total resource number from the observation model (`observation_array` also holds this information in a different way depending on `observe_type`)
- `sim_new$LAND`: The landscape on which interactions occur, which is stored as a 3D array with `land_dim_1` rows, `land_dim_2` columns, and 3 layers. Layer 1 (`sim_new$LAND[,,1]`) is not currently used in default submodels, but could be used to store values that affect resources and agents. Layer 2 (`sim_new$LAND[,,2]`) stores crop yield from a cell, and layer 3 (`sim_new$LAND[,,3]`) stores the owner of the cell (value corresponds to the agent's ID).
- `sim_new$manage_vector`: The cost of each action as set by the manager. For even more fine-tuning, individual costs for the actions of each agent can be set for each user in `sim_new$manager_array`.
- `sim_new$user_vector`: The total number of actions performed by each user. A more detailed breakdown of actions by individual users is held in `sim_new$user_array`.

Next, we show how to adjust the landscape to manually set land ownership in `gmse_apply`.

@@

By default, all farmers in GMSE are allocated the same number of landscape cells, which are simply placed in order of the farmer's ID. Public land is produced by placing landscape cells that are technically owned by the manager, and therefore have landscape cell values of 1. The image below shows this landscape for the eight farmers from `sim_old`.

@@

```{r, echo = TRUE, eval = TRUE}
sim_old  <- gmse_apply(get_res = "Full", remove_pr = 0.122, lambda = 0.275,
                       res_d4eath_K = 93870, RESOURCE_ini = 35000, 
                       manage_target = 70000, res_death_type = 3,
                       manager_budget = 10000, user_budget = 10000,
                       public_land = 0.4, stakeholders = 8, res_consume = 0.02, 
                       res_birth_K = 200000, land_ownership = TRUE,
                       observe_type = 3, agent_view = 1, converge_crit = 0.01,
                       ga_mingen = 200);
```

@@

Perhaps we want to assume that farmers have different budgets, which are correlated in some way to the number of landscape cells that they own. Custom user budgets can be set by manipulating `sim_old$AGENTS`, the last column of which (column 17) holds the budget for each user. Agent IDs (as stored on the landscape above) correspond to rows of `sim_old$AGENTS`, so individual budgets can be directly input as desired. We can do this manually (e.g., `sim_old$AGENTS[2, 17] <- 4000`), or, alternatively, if farmer budget positively correlates to landscape owned, we can use a loop to input values as below.

@@

```{r, echo = FALSE}
library(pander);
budget_table <- rbind(sim_old$AGENTS[,1], sim_old$AGENTS[,17]);
row.names(budget_table) <- c("ID", "Budget");
pander(budget_table);
```

@@

Lastly, we consider a more nuanced change to simulations, in which the rules for movement of resources are modified to account for density-dependence. Assume that geese tend to avoid aggregating, such that if a goose is located on the same cell as too many other geese, then it will move at the start of a time step. Programming this movement rule can be accomplished by creating a new function to apply to the resource data array `sim_old$resource_array`. Below, a custom function is defined that causes a goose to move up to 5 cells in any direction if it finds itself on a cell with more than 10 other geese. As with default GMSE simulations, movement is based on a torus landscape.

```{r}
avoid_aggregation <- function(goose_table, land_dim_1 = 100, land_dim_2 = 100){
    goose_number <- dim(goose_table)[1]    # How many geese are there?
    for(goose in 1:goose_number){          # Loop through all rows of geese
        x_loc  <- goose_table[goose, 5];
        y_loc  <- goose_table[goose, 6];
        shared <- sum(goose_table[,5] == x_loc & goose_table[,6] == y_loc);
        if(shared > 10){
            new_x <- x_loc + sample(x = -5:5, size = 1);
            new_y <- y_loc + sample(x = -5:5, size = 1);
            if(new_x < 0){ # The 'if' statements below apply the torus
                new_x <- land_dim_1 + new_x;
            }
            if(new_x >= land_dim_1){
                new_x <- new_x - land_dim_1;
            }
            if(new_y < 0){
                new_y <- land_dim_2 + new_x;
            }
            if(new_y >= land_dim_2){
                new_y <- new_y - land_dim_2;
            }
            goose_table[goose, 5] <- new_x;
            goose_table[goose, 6] <- new_y;
        }
    }
    return(goose_table);
}
```

With the above function written, we can apply the new movement rule along with our [custom farm placement](#land) and [custom farmer budgets](#budgets) to the simulation of goose population dynamics.

Simulation with custom farms, budgets, and goose movement
================================================================================

Below shows an example of `gmse_apply` with custom landscapes, farmer budgets, and density-dependent goose movement rules.

```{r, echo = TRUE, eval = TRUE}
# First initialise a simulation
sim_old  <- gmse_apply(get_res = "Full", remove_pr = 0.122, lambda = 0.275,
                       res_death_K = 93870, RESOURCE_ini = 35000, 
                       manage_target = 70000, res_death_type = 3,
                       manager_budget = 10000, user_budget = 10000,
                       public_land = 0.4, stakeholders = 8, res_consume = 0.02, 
                       res_birth_K = 200000, land_ownership = TRUE,
                       observe_type = 3, agent_view = 1, converge_crit = 0.01,
                       ga_mingen = 200, res_move_type = 0);
# By setting `res_move_type = 0`, no resource movement will occur in gmse_apply
# Adjust the landscape ownership below
sim_old$LAND[1:20,   1:20,  3] <- 2; 
sim_old$LAND[1:20,  21:40,  3] <- 3;
sim_old$LAND[1:20,  41:60,  3] <- 4;
sim_old$LAND[1:20,  61:80,  3] <- 5;
sim_old$LAND[1:20,  81:100, 3] <- 6;
sim_old$LAND[21:40,  1:50,  3] <- 7;
sim_old$LAND[21:40, 51:100, 3] <- 8;
sim_old$LAND[41:60,  1:100, 3] <- 9;
sim_old$LAND[61:100, 1:100, 3] <- 1;
# Change the budgets of each farmer based on the land they own
for(ID in 2:9){
    cells_owned            <- sum(sim_old$LAND[,,3] == ID);
    sim_old$AGENTS[ID, 17] <- 10 * cells_owned;
}
# Begin simulating time steps for the system
sim_sum_2 <- matrix(data = NA, nrow = 30, ncol = 5);
for(time_step in 1:30){
    # Apply the new movement rules at the beginning of the loop
    sim_old$resource_array  <- avoid_aggregation(sim_old$resource_array);
    # Next, move on to simulate (old_list remembers that res_move_type = 0)
    sim_new                 <- gmse_apply(get_res = "Full", old_list = sim_old);
    sim_sum_2[time_step, 1] <- time_step;
    sim_sum_2[time_step, 2] <- sim_new$basic_output$resource_results[1];
    sim_sum_2[time_step, 3] <- sim_new$basic_output$observation_results[1];
    sim_sum_2[time_step, 4] <- sim_new$basic_output$manager_results[3];
    sim_sum_2[time_step, 5] <- sum(sim_new$basic_output$user_results[,3]);
    sim_old                 <- sim_new;
}
colnames(sim_sum_2) <- c("Time", "Pop_size", "Pop_est", "Cull_cost", 
                         "Cull_count");
print(sim_sum_2);
```

Conclusions
================================================================================

In this example, we showed how the built-in resource, observation, manager, and user sub-models can be customised by manipulating the data within the data structures that they use. The goal was to show how software users can work with these existing sub-models and data structures to customise GMSE simulations. Software users seeking even greater flexibility (e.g., replacing an entire built-in submodel with a custom submodel) should refer to the [Supporting Information 2](https://bradduthie.github.io/gmse_SI2.pdf) that introduces `gmse_apply` more generally. Future versions of GMSE are likely to expand on the built-in options available for simulation; requests for such expansions, or contributions, can be submitted to [GitHub](https://github.com/bradduthie/gmse/issues).
